---
title: "download_montreal_presence_absence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{download_montreal_presence_absence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Downloading bird presence absence using R Atlas

## Functions description

Bird presence / absence are generated through observations datasets that are
qualified as `exhaustive`, meaning datasets from which we can deduce an absence
for a species at a specific date - time - location combination (defined as 
sampling points).

Two distinct functionalites are available in ratlas for that purpose.

* `list_bird_quebec_taxa_ref()` returns a dataframe containining all taxons
observed within quebec`s territory, for any rank, taxon reference source or
validity of taxons. The returned objects provide also general statistics on
occurrence of species.

* `get_bird_presence_absence(taxa_name)` returns a dataframe of the same amount
of sampling points, for which the attribute `occurrence` is `TRUE` if the taxon 
has been observed at the sampling points, or `FALSE` otherwise. Other attributes
related to position, date, time, taxon and observation datasets are included.

The equivalent functions for Montreal's territory are respectively
`list_bird_mtl_taxa_ref()` and `get_mtl_bird_presence_absence(taxa_name)` with
the same usage.

## Usage

*Remarks about libraries* In the folloiwing examples, Data.frame manipulation is
done through `dplyr`, geographic type handling is done through `sf` and the
display of geographic data is done with `leaflet`.

```{r setup, warning=FALSE}
library(ratlas)
library(dplyr)
library(sf)
library(leaflet)
library(knitr)
```
```{r, include = FALSE}
kable <- function(data) {
  knitr::kable(data) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
}
```

# List observed bird taxons

A single call will list all observed taxons within quebec's territory. However,
additionnal parameters may be added to filter `taxa` records through their
rank and their validity.

*Example : List only valid species*
```{r}
species <- list_mtl_bird_taxa(rank = 'species', valid = TRUE)
head(species) %>% kable()
```

There are multiple valid record for a single species since we are returning a
record for each taxonomic reference backbone, since some of them might disagree
on validity of taxon for a single organism. A unique list of organism name
may be generated from the taxonomic `source_name` or using distinct
`scientific_name`.

```{r, render: knitr::kable}
# Unique organism using gbif taxonomic backbone
species %>% dplyr::filter("source_name" == "GBIF Backbone Taxonomy")

# Unique organism using distinct `scientific_name`
species <- species %>% dplyr::distinct(scientific_name, .keep_all = TRUE)
```

A general understanding of the frequency and characteristics of occurrences
may be understood through the values of `max_yearly_count`, `avg_yearly_count`,
`min_yearly_count`, `total_count`, `min_year_obs` and `max_year_obs`.
We thus can order the species by the `total_count` of occurrences in descending
order and inspect the 10 most observed species.

```{r}
species %>% arrange(-total_count) %>% head(10) %>% kable()
```

# Get presence/absence data

We can obtain a data.frame containing the presence and absence of a taxon
using the function `get_mtl_bird_presence_absence()`. The function downloads and
returns a dataframe containing presence and absence (values TRUE or FALSE) with
a single record per row for the bird species specified by `taxa_name`.

The absence are infered from all sampling points related to birds observed on
Montreal territory whose observations were obtained from datasets defined as
exhaustive, from the `eBird` and `Atlas des oiseaux nicheurs du Quebec`
datasets.

```{r}
occurences <- get_mtl_bird_presence_absence(taxa_name = "Picoides pubescens")
head(occurences) %>% kable()
```

The `geom` attributes provide a Well Known Text (WKT) of the position of
occurrence and may by put into a `sf` simple feature data.frame for manipulation
and easy mapping. Here, we filter only presence where `occurrence` is `TRUE` 
since year 2015 and plot them on a map.

```{r}
occurrences <- st_as_sf(occurences, wkt = "geom")
map <- leaflet() %>%
  addTiles() %>%
  addMarkers(data = filter(occurrences, occurrence == TRUE, year_obs >= 2015),
    clusterOptions = markerClusterOptions())
map
```
