---
title: "Download observations from Atlas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{download-obs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Dowloading observations

## Get started quickly : Get *Canis Lupus* observations from Atlas

Observations within Atlas may be downloaded using the `get_observations`
functions. To filter our request to the dataset, we will use the function
`get_taxa()` to obtain first the `Canis Lupus` taxa data. We then use it to
as a filter to get observations and then display the first 10 records.

```{r}
library(ratlas)
taxa <- get_taxa(scientific_name = "Canis lupus")
obs <- get_observations(id_taxa = taxa$id_taxa_obs)
head(obs, 10)
```

RAtlas functions expose Atlas data tables and related attributes. We can access
related records storing information about the `datasets` and measured
`variables` and filter them using the `id_datasets` and `id_variables` values
for our observations.

We can finally join the taxa, dataset and variables attributes to the table
using the `dplyr` functions and `%>%` pipe operation.

```{r}
library(magrittr) # To use pipe operator %>%

# Get data from Atlas tables
var <- get_gen('variables', id=obs$id_variables)
datasets <- get_datasets(obs$id_datasets)

# Rename columns with prefix before join
taxa <- dplyr::rename_with(taxa, ~ paste0("taxa.", .x))
var <- dplyr::rename_with(var, ~ paste0("var.", .x))
datasets <- dplyr::rename_with(datasets, ~ paste0("ds.", .x))

# Join using `left_join` function from `dplyr`
obs <- obs %>%
  dplyr::left_join(taxa, by = c("id_taxa" = "taxa.id")) %>%
  dplyr::left_join(var, by = c("id_variables" = "var.id")) %>%
  dplyr::left_join(datasets, by = c("id_datasets" = "ds.id"))

# Display first record as a vertical list
print(t(obs[1, ]))
```

## Filter observation request using any table attributes

Any request may be filtered using any data attributes used as a data column in
Atlas. For example, we will filter the `original_source` column to get only
datasets related to the `eBird` datasets. We will the use the id values of those
datasets to return only `observations` with the specified `id_datasets` and
for the taxa associated to the Snowy owl *Bubo scandiacus*.

```{r}
# Get dataset and taxa records for which we will filter observations
ebird <- get_datasets(original_source = "eBird") %>%
  dplyr::rename_with(~ paste0("ds.", .x))
taxa <- get_taxa(scientific_name = "Bubo scandiacus") %>%
  dplyr::rename_with(~ paste0("taxa.", .x))

# Filter observation request through `id_taxa` and `id_datasets` and year values
obs <- get_observations(
  id_datasets = ebird$ds.id,
  id_taxa = taxa$taxa.id,
  year = 2018:2021)

# Join using taxa, datasets, and variables values
var <- get_gen('variables', id=unique(obs$id_variables)) %>%
  dplyr::rename_with( ~ paste0("var.", .x))

obs <- obs %>%
  dplyr::left_join(taxa, by = c("id_taxa" = "taxa.id")) %>%
  dplyr::left_join(var, by = c("id_variables" = "var.id")) %>%
  dplyr::left_join(ebird, by = c("id_datasets" = "ds.id"))

print(glue::glue(
  "Found {nrow(obs)} observations between {min(obs$year_obs)} ",
  "and {max(obs$year_obs)}")
  )

# Print map of observations using the package `leaflet`

# install.packages("leaflet")
library(leaflet)
coord <- head(t(sapply(obs$geom.coordinates, as.list)), 100) # Only 100 records shown
colnames(coord) <- c("lng", "lat")

map <- leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng = as.numeric(coord[, "lng"]), lat = as.numeric(coord[, "lat"]))
map
```

