---
title: "Download observations from Atlas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{download-obs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Dowloading observations

## Get started quickly : Get *Canis Lupus* observations from Atlas

Observations within Atlas may be downloaded using the `get_observations`
functions. To filter our request to the dataset, we will use the function
`get_taxa()` to obtain first the `Canis Lupus` taxa data. We then use it 
as a filter to get observations and then display the first 10 records.

```{r}
library(ratlas)
taxa <- get_taxa(scientific_name = "Canis lupus")
obs <- get_observations(id_taxa = taxa$id_taxa_obs)
head(obs, 10)
```

RAtlas functions expose Atlas data tables and related attributes.
The observations table returned by the `get_observations` function 
already returns informations on the related dataset and taxonomy of 
an observations.

```{r}
# Display first record as a vertical list
print(t(obs[1, ]))
```

## Filter observation request using any table attributes

Any request may be filtered using any data attributes used as a data column in
Atlas. For example, we will filter the `original_source` column to get only
datasets related to the `Placettes-échantillons permanentes` datasets.
We will then use the id values of those
datasets to return only `observations` with the specified `id_datasets` and
for the taxa associated to the sugar maple *Acer saccharum*.

Different tables can be joined using the base pipe `|>` and `dplyr` functions.

```{r}
# Get dataset and taxa records for which we will filter observations
pep <- get_datasets(original_source = "Placettes-échantillons permanentes") |>
  dplyr::rename_with(~ paste0("ds.", .x))
taxa <- get_taxa(scientific_name = "Acer saccharum") |>
  dplyr::rename_with(~ paste0("taxa.", .x))

# Filter observation request through `id_taxa` and `id_datasets` and year values
obs <- get_observations(
  id_datasets = pep$ds.id,
  id_taxa = taxa$taxa.id_taxa_obs,
  year = 2018:2021)

print(glue::glue(
  "Found {nrow(obs)} observations between {min(obs$year_obs)} ",
  "and {max(obs$year_obs)}")
  )

# Print map of observations using the package `leaflet`

# install.packages("leaflet")
library(leaflet)
coord <- head(t(sapply(obs$geometry, as.list)), 100) # Only 100 records shown
colnames(coord) <- c("lng", "lat")

map <- leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng = as.numeric(coord[, "lng"]), lat = as.numeric(coord[, "lat"]))
map
```

